A general and efficient representation of Ancestral Recombination Graphs

### 2022-09-01 paper.tex
Tweaks

The main reasons for wanting precise estimates of all recombination
events are to either compute a likelihood of the inferred ARG
under a statistical model, or to study the recombinants in
detail~\citet{rasmussen2022espalier}.
% Explain that we're really interested in the details of recombination
% itself with a small number of samples.
In the latter case, we might
have a small number of samples~\citep{guo2022recombination}.

@@

% If we want to use the coalescent, then really the assumptions
% need to make sense.
% n << Ne is a silly assumption in today's datasets
A core assumption of the coalescent is that the sample size $n$
is much less than the effective population size, $N_e$.
% This
% assumption ensures that multiple events (i.e., simultaneous
% coalescence and recombination) don't occur at the same
% time, which is reflected in the eARG definitions.
This once-reasonable assumption has been overtaken by
the data we have available.
% This can be tightened up a bit, I think. We may need to drop a
% few references.
Several human datasets now consist of hundreds of thousands of
genomes~\citep{bycroft2018genome,karczewski2020mutational,tanjo2021practical},
and so sample size is substantially \emph{larger} than $N_e$
(often assumed to be $10^4$ in humans).
Agricultural datasets are even more extreme.
For example,
the US dairy cattle database alone currently comprises more than 6 million
animals with SNP array
genotypes\footnote{\url{https://queries.uscdcb.com/Genotype/counts.html}},
while the effective population size in modern dairy cattle breeds is
less than 100 and decreasing~\citep{MacLeod2013,Makanjuola2020}.
An extreme sign of these breeding practices is that there are only two ancestral
Y-chromosome linages present in today's US Holstein dairy breed~\citep{Yue2015}.
Simlarly, the 1000 bull genomes project~\citep{hayes20191000}
comprises close to 7000 genomes, which are part of multi-generation pedigrees
with millions of animals and extensive SNP array genotype and phenotype
data \citep[e.g.][]{Cesarani2022}.
In another example, genome and SNP array data for 440,610 individuals within
7 multi-generation pedigrees~\citep{whalen2018,Johnsson2021,Ros-Freixedes2020}
were combined to infer recombination in pigs~\citep{RosFreixedes2022}.
Effective population size in modern pig breeds is also less than 100 due to
intense selection and directed reproduction \citep{Hall2016,Porcnic2016}.
Another core assumption of the coalescent model is that the genome (or
at least the region under study) is short enough that the number of extant
lineages remains much smaller than $N_e$ at all times. Whole
genome sequences have been available for model organisms
for over a decade now, % True? Citation?
and indeed complete chromosome-level assemblies are possible
in humans~\citep{miga2020telomere}.
Projects are under way to obtain high-quality assemblies
for all eukaryotic species in Britain and Ireland~\citep{darwin2022sequence}
and ultimately worldwide~\citep{lewin2022earth}.
% Clearly multiple recombinations (and other complex events) will
% occur in the ancestry of such datasets, and we should not be
% limited in what we can infer by the data structure that we
% use to represent the results.

While the coalescent as a model can be suprisingly robust to
basic violations of the underlying assumptions~\citep{
wakeley2012gene,bhaskar2014distortion,nelson2020accounting}, [something
something]

Such datasets cannot be reasonably be described by
the coalescent as a statical model, and so more non-parametric
methods like \tsinfer\ and \relate\ would seem more appropriate.
[More]

% Is it practical to estimate full ARGs?
[Not particularly \citep{shipilina2022origin}]

>>>>

% What is the goal of having fully precise ARGs?
% 1) studying the actual recombination events;
% 2) evaluating the likelihoods
The main reasons for wanting precise estimates of all recombination
events are to either compute a likelihood of the inferred ARG
under a statistical model, or to study the recombinants in
detail~\citep{rasmussen2022espalier}.
% Explain that we're really interested in the details of recombination
% itself with a small number of samples.
In the latter case, we might
have a small number of samples~\citep{guo2022recombination}.

@@

% If we want to use the coalescent, then really the assumptions
% need to make sense.
% n << Ne is a silly assumption in today's datasets
A core assumption of the coalescent is that the sample size $n$
is much less than the effective population size, $N_e$.
Several human datasets now consist of hundreds of thousands of
genomes~\citep{bycroft2018genome,karczewski2020mutational,tanjo2021practical},
and so sample size is substantially \emph{larger} than $N_e$
(often assumed to be $10^4$ in humans).
Agricultural datasets are even more extreme.
For example,
the US dairy cattle database alone currently comprises more than 6 million
animals with SNP array
genotypes\footnote{\url{https://queries.uscdcb.com/Genotype/counts.html}},
while the effective population size in modern dairy cattle breeds is
less than 100 and decreasing~\citep{MacLeod2013,Makanjuola2020}.
An extreme sign of these breeding practices is that there are only two ancestral
Y-chromosome linages present in today's US Holstein dairy breed~\citep{Yue2015}.
Simlarly, the 1000 bull genomes project~\citep{hayes20191000}
comprises close to 7000 genomes, which are part of multi-generation pedigrees
with millions of animals and extensive SNP array genotype and phenotype
data \citep[e.g.][]{Cesarani2022}.
In another example, genome and SNP array data for 440,610 individuals within
7 multi-generation pedigrees~\citep{whalen2018,Johnsson2021,Ros-Freixedes2020}
were combined to infer recombination in pigs~\citep{RosFreixedes2022}.
Effective population size in modern pig breeds is also less than 100 due to
intense selection and directed reproduction \citep{Hall2016,Porcnic2016}.
Another core assumption of the coalescent model is that the genome (or
at least the region under study) is short enough that the number of extant
lineages remains much smaller than $N_e$ at all times. Whole
genome sequences have been available for model organisms
for over a decade now, % True? Citation?
and indeed complete chromosome-level assemblies are possible
in humans~\citep{miga2020telomere}.
Projects are under way to obtain high-quality assemblies
for all eukaryotic species in Britain and Ireland~\citep{darwin2022sequence}
and ultimately worldwide~\citep{lewin2022earth}.

While the coalescent as a model can be suprisingly robust to
basic violations of the underlying assumptions~\citep{
wakeley2012gene,bhaskar2014distortion,nelson2020accounting}
it is reasonable to question the validity of the model in
representing datasets such as UK Biobank.
[Also mention somehow that $n\ll N_e$ is only one problem -
we're also ignoring the myriad other evolutionary forces.]
Such datasets cannot be reasonably be described by
the coalescent as a statical model, and so more non-parametric
methods like \tsinfer\ and \relate\ would seem more appropriate.
[More]

% Is it practical to estimate full ARGs?
% [Not particularly \citep{shipilina2022origin} - but maybe ARGneedle
% has done this for UKB? Need to check]
